---
title: "Final_Kmodes"
output: html_document
date: "2023-11-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
library(klaR)
library(cluster)

data <- read.csv("h1n1_data.csv")
data <- data[, -1]
data <- data[, !names(data) %in% c("seasonal_vaccine", "h1n1_vaccine", "vaccination_status")]


data[] <- lapply(data, function(x) {
  if (is.numeric(x)) {
    factor(ifelse(is.na(x), "Missing", as.character(x)))
  } else {
    factor(ifelse(is.na(x), "Missing", x))
  }
})


Csilhouette <- function(data, k) {
  cluster_assignment <- kmodes(data, k)$cluster
  silhouette_width <- silhouette(cluster_assignment, dist(data, method = "binary"))
  mean(silhouette_width[, 3])
}


k_values <- 3:10
silhouette_values <- sapply(k_values, function(k) Csilhouette(data, k))
silhouette_values

best_k <- k_values[which.max(silhouette_values)]
best_k
```
```{r}
kmodes_result <- kmodes(data, 3)
kmodes_result$size
kmodes_result$mode
```
```{r}
modes <- kmodes_result$mode

common_variables <- which(apply(modes, 2, function(x) length(unique(x)) == 1))

common_modes <- modes[1, common_variables]
names(common_modes)
```

## Report
Our objective in this study was to uncover patterns in a dataset centered around H1N1 awareness and vaccination behaviors. We opted for k-modes clustering, a method well-suited for categorical data, to discern distinct groups within the population based on their responses and characteristics.

The initial dataset was rich in categorical variables, spanning demographic details and behavioral responses linked to H1N1. In preparing for clustering, we undertook several modifications:

1. We removed the 'seasonal_vaccine', 'h1n1_vaccine', and 'vaccination_status' columns to avoid any bias in the formation of clusters.
2. The first column, appearing to be an identifier, was excluded.
3. We converted all numerical columns to categorical formats, aligning them with the requirements of k-modes clustering.

Utilizing the klaR package in R, we performed k-modes clustering across a range of potential cluster numbers, from 3 to 10. The optimal number of clusters was determined based on the average silhouette width, a measure of how well an object fits within its cluster relative to other clusters. 

Initially, we considered a cost function analysis for selecting the optimal number of clusters. This function measures the dissimilarities of objects to their nearest cluster modes. However, this approach was set aside due to:

1. Computational Complexity: The computation of the cost function across various k values proved to be highly demanding, especially considering the size and complexity of our dataset.
2. Ambiguity in Interpretation: We found that the cost function tended to decrease with more clusters, lacking a clear 'elbow' point, which made selecting the optimal k somewhat subjective and potentially misleading.

Consequently, we selected k=3 based on silhouette scores, leading to the identification of three distinct groups in our dataset. Notably, 17 variables exhibited the same mode across all clusters:

[h1n1_knowledge, behavioral_antiviral_meds, behavioral_avoidance, behavioral_face_mask, behavioral_wash_hands, behavioral_large_gatherings, behavioral_outside_home, behavioral_touch_face, chronic_med_condition, child_under_6_months, health_worker, health_insurance, opinion_seas_sick_from_vacc, race, sex, rent_or_own, hhs_geo_region, household_children]

The uniformity in these variables across diverse groups reveals that specific beliefs, behaviors, and demographic traits are common among various population segments:

1. Knowledge and Behavioral Practices: The consistency in variables such as 'h1n1_knowledge', 'behavioral_antiviral_meds', and 'behavioral_avoidance' across all clusters indicates a widespread awareness and adoption of similar health practices, irrespective of the cluster. This trend is a positive sign from a public health standpoint.

2. Health Conditions and Worker Status: Uniformity in 'chronic_med_condition', 'health_worker', and 'child_under_6_months' suggests these health-related characteristics are evenly distributed across the population, possibly reflecting broader population trends or effective health communication.

3. Socio-Demographic Factors: The consistent distribution of 'race', 'sex', 'rent_or_own', and 'hhs_geo_region' across clusters suggests that these demographic factors do not significantly influence cluster formation. This indicates a uniform impact or response to H1N1 that crosses demographic lines.

4. Health Perceptions and Insurance: The general agreement in views on 'health_insurance' and 'opinion_seas_sick_from_vacc' across clusters may reflect a shared public sentiment about health insurance and vaccine perceptions.


```

